
\documentclass{ximera}

\input{../../preamble.tex}

\author{David Kish}
\license{Creative Commons Attribution-ShareAlike 4.0 International License}



\begin{document}
The following exponential equation models how a particular medicine degrades in a patient's body over time. In this model, $x$ represents the number of hours that the medicine has been in the patient and $y$ represents the amount of medicine in milligrams left in the patient after $x$ hours.
\[
y=300(0.75)^x
\]
Use the equation to help answer the questions below.

\begin{exercise}
Which of the following are correct interpretations of $.075$ from the model?
  \begin{selectAll}
    \choice{The medicine in the patient increases by $75\%$ each hour.}
    \choice[correct]{$75\%$ of the medicine remains in the patient after each hour.}
    \choice{The medicine in the patient decreases by $0.75$ milligrams every hour.}
\choice{The medicine in the patient increases by $0.75$ milligrams every hour}
    \choice[correct]{The medicine in the patient degrades by $25\%$ each hour.}
  \end{selectAll}
\end{exercise}
\begin{exercise}
What is the value of $y$ when $x=0$?\\
$y=\answer{300}$ \wordChoice{ \choice[correct]{milligrams} \choice{hours}}
\end{exercise}
\begin{exercise}
Which of the following are correct about the answer to the previous question?
  \begin{selectAll}
    \choice[correct]{The answer represents the initial dose of medicine given to the patient.}
    \choice{The answer represents when there is no medicine left in the patient}
    \choice[correct]{The answer represents the amount of medicine in the patient after $0$ hours.}
\choice{The answer represents the amount of time it takes for all of the medicine to leave the patient's body}
    \choice{The answer represents the amount of medicine in the patients body at the end of the day.}
  \end{selectAll}
\end{exercise}
\begin{exercise}
What is the remaining drug concentration after a day has passed?\\
Please enter your answer as a calculator ready response.\\
$y=\answer{300(0.75)^{24}}$milligrams.
\end{exercise}
\begin{exercise}
We want to know how long it takes for the drug concentration to be cut in half. How would we setup the model to figure this out?\\
$\answer{150}=300(0.75)^{\answer{x}}$
\end{exercise}
\end{document}