
\documentclass{ximera}

\input{../../preamble.tex}

\author{Debi Stout}
\license{Creative Commons Attribution-ShareAlike 4.0 International License}



\begin{document}
Nancy wants to put a nice patio in her backyard. She decides she’d like it to be
concrete, and starts putting in the research to see what she’ll need. Nancy would like
her patio to be 12 feet 8 inches wide (the width between her flower beds) and 10 feet
out. Here’s some information she found online to help her with this project:
\begin{itemize}
\item Concrete should probably be poured about 4 inches thick
\item 80 lb bags of concrete are available at Local Hardware Store for \$4.65.
\item Those concrete packages say that with adding 6 pints of water to each 80 lb bag, she can expect to get about 0.60 cubic feet of concrete.
\end{itemize}

\begin{exercise}
How many cubic feet of concrete will Nancy need? \\
Give the exact calculator-ready answer.  $ \answer{42+\frac{2}{9}} \text{ft}^3$ \\
Now, round your answer up to the nearest integer to be on the safe side.  Nancy wants $ \answer{43}$ \calcHW $\text{ft}^3$ of concrete. \\
\begin{exercise}
How many 80 lb bags of concrete should Nancy buy, assuming she wants 43 cubic feet of concrete?\\
Give the exact calculator-ready answer.  $ \answer{71+\frac{2}{3}}$ bags of concrete.\\
Now, round your answer up to the nearest integer.  Nancy can't buy a partial bag!  Nancy wants $ \answer{72}$ \calcHW  bags of concrete. \\
\begin{exercise}
Assuming Nancy wants to mix up all 72 bags of concrete, how much water will she need? \\
She will need $\answer{54}$ gallons of water.\\
\begin{exercise}
How much will all those bags of concrete cost Nancy?  Assume she is buying them in Columbus and include the 7.5\% sales tax charged in Columbus in her total cost.  If necessary, round your answer to the nearest dollars and cents.\\
Nancy will pay \$ $\answer{359.91}$ \calcHW .
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{document}