\documentclass{ximera}

\input{../../preamble.tex}

\author{David Kish}




\begin{document}
You are encouraged to solve this problem without the use of a calcutor. A ``calculator ready'' response can be submitted in the answer box.

You may use the following unit conversions while completing the problems below:
\begin{itemize}
\item 3 teaspoons = 1 Tablespoon
\item 4  Tablespoons = $\frac{1}{4}$ cup
\item 4 cups = 1 quart
\item 2 pints = 1 quart
\item 4 quarts = 1 gallon
\end{itemize}

\begin{exercise}
 How many tablespoons are in a gallon? 
 $\answer{256}$ tablespoons $= 1$ gallon
 \end{exercise}
  \begin{exercise}
 How many cups of gasoline could fit into a 15 gallon tank?
 $\answer{240}$ cups $= 15$ gallons
 \end{exercise}
  \begin{exercise}
 How many teaspoons are in a pint?
 $\answer{96}$ teaspoons $= 1$ pint
 \end{exercise}
  \begin{exercise}
 How many gallons are in a pint?
 $\answer{\frac{1}{8}}$ gallons $=1 $ pint
 \end{exercise}

For the following problems using the metric system, you may use the Conversion factors available here: \link[ Metric System and Equivalents]{https://www.nwcg.gov/course/ffm/conversions/table-23-metric-system-and-equivalents}

  \begin{exercise}
 How many centimeters are in a kilometer? 
 $\answer{100000}$ cm $= 1$ km
 \end{exercise}
   \begin{exercise}
 $53$ decaliters is how many milliliters?
 $\answer{530000}$ mL $= 53$ daL
 \end{exercise}
   \begin{exercise}
 How many decigrams are in a hectogram? 
$ \answer{1000}$ dg$ = 1$ hg
 \end{exercise}
   \begin{exercise}
 How many millimeters are in a decimeter?
 $\answer{100}$ mm $= 1$ dm
 \end{exercise}


\end{document}
