\documentclass{ximera}

\input{../../preamble.tex}

\author{David Kish}




\begin{document}
   \begin{exercise}
Nancy wants to put a nice patio in her backyard.  She decides she’d like it to be concrete, and starts putting in the research to see what she’ll need.  Nancy would like her patio to be 12 feet 8 inches wide (the width between her flower beds) and 10 feet out.  Here’s some information she found online to help her with this project:
\begin{itemize}
\item Concrete should probably be poured about 4 inches thick.
\item 80 lb bags of concrete are available at Local Hardware Store for $4.65.
\item Those concrete packages say that with adding 6 pints of water to each 80 lb bag, she can expect to get about 4.5 gallons of concrete.
\end{itemize}
\begin{enumerate} 
\item  How many cubic feet of concrete will Nancy need?
$\answer{}$ cubic feet
\item How much is that volume in gallons?
(1 cubic foot = 7.48 gallons)
\item How many 80 lb bags of concrete should Nancy buy?

\item How much will buying all of these bags of concrete cost her? 
 
\item How many pints of water will she need for all of her concrete?
\end{enumerate}

 You have a coupon for $30\%$ off and want to buy an item that is $\$48.98$. 

How much money will you save? $\$\answer{14.69}$

How much money will you pay? $\$\answer{34.29}$
 \end{exercise}
%\begin{multipleChoice}
%\choice{constant function.}
%\choice{positive function. }
%\choice{negative function.}
%\choice{position function.}
%\choice[correct]{velocity function.}
%\end{multipleChoice}


\end{document}