\documentclass{ximera}

\input{../../preamble.tex}

\author{Kenneth Berglund}




\begin{document}
In this exercise, we will rewrite the following quadratic in vertex form $a(x - h)^2 + k$, using the method of completing the square:
\[
f(x)=4x^2-24x+11.
\]
\begin{exercise}

Factoring the leading coefficient out of the $x^2$ and $x$ terms yields
$$
\answer{4}(x^2 + \answer{-6}x) + \answer{11}.
$$

\begin{exercise}
Notice that we can't rewrite $x^2 - 6x$ as a square of a linear polynomial. We need to add a constant term in order to make $x^2 - 6x$ into a square. 

We want to find numbers a number $h$ such that $(x - h)^2 = x^2 - 6x + h^2$.  

$$
h = \answer{3} \text{ and } h^2 = \answer{9}
$$
\begin{hint}
Try multiplying out $(x - h)^2$ and matching the coefficients of each term with the coefficients of the corresponding term in $x^2 - 6x + h^2$. 
\end{hint}


\begin{exercise}
Modify the quadratic to introduce the needed $h^2$ term.
$$
4(x^2 - 6x + \answer{9} - \answer{9}) + 11
$$ 

Use positive numbers in your answer.

\begin{hint}
Remember that if you add a term to an expression, to preserve equality, you must subtract the same term.
\end{hint}

\begin{exercise}
Now, regroup and distribute the 4 so you end up with an expression in the form below.
$$
4(x^2-6x+9) + \answer{-25}
$$ 

\begin{exercise}
Now rewrite $x^2 - 6x + 9$ as a square of a linear polynomial to finish completing the square!
$$
4(x - \answer{3})^2 -25
$$ 
\begin{hint}
What square is equal to $x^2 - 6x + 9$? We found it before!
\end{hint}

\begin{exercise}
As a bonus, here's what the manipulation would look like when done all at once:
\begin{align*}
4x^2 - 24x + 9 & = 4(x^2 - 6x) + 11 \\
& = 4(x^2 - 6x + 9 - 9) + 11 \\
& = 4(x^2-6x+9)-36 + 11\\
& = 4(x^2-6x+9)-25\\
& = 4(x - 3)^2 - 25
\end{align*}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}
\end{exercise}



\end{document}