\documentclass{ximera}

\input{../../preamble.tex}

\author{Elizabeth Campolongo}
\acknowledgement{https://www.stitz-zeager.com/szprecalculus07042013.pdf}

\begin{document}
\begin{exercise}
For the following word problems, determine what type of function is needed to solve them (e.g., linear, quadratic, cubic, logarithmic, exponential, etc.) and give the function. 

Mary has decided to take up alpaca farming to make home-spun yarn, thus feeding her other hobby of sustainable knitting. She just inherited a large parcel of land in Ashtabula, which is perfect for her to make this dream come true! Her funds are still limited, so she can afford only enough fencing material to build a rectangular pasture with 200 linear feet of fence. She will construct it adjacent to a conveniently placed stream (removing the requirement for fencing on this side). 
%ex2.3.4 in stitz-zeager

She wants to maximize the area of the pasture. What type of function will she need to do this calculation?\\
\wordChoice{\choice{linear}\choice[correct]{quadratic}\choice{cubic}\choice{logarithmic}\choice{exponential}}
\begin{exercise}
She will need a quadratic equation to maximize the Area. If we let $\ell$ denote the length adjacent to the river and $w$ denote the width of the pasture, the formula for the area in terms of width is
$$A(w) = \answer{-2}w^2 + \answer{200w}.$$

\begin{exercise} %ex20 of same section
Mary starts to worry that leaving the alpaca along the stream could be dangerous with the storms that have been rolling in recently, and she decides to construct a freestanding pasture. She still only has 200 feet worth of fencing; what is the new function to maximize the area in terms of width?\\
$A(w) = \answer{-}w^2 + \answer{100w}$.
\end{exercise}
\end{exercise}


\end{exercise}
\end{document}