\documentclass{ximera}

\input{../preamble.tex}
\author{Ivo Terek, Elizabeth Miller}
\license{Creative Commons Attribution-ShareAlike 4.0 International License}
\acknowledgement{}
%Source: Stitz-Zeager
\title{End Behavior of Rational Functions}

\begin{document}
\licenseSZ
\begin{abstract}
\end{abstract}
\maketitle

As we have with previous functions we have studied, let's look at the end behavior of rational functions.  Recall that we noticed that the function $f(x)=\frac{1}{x}$ approached 0 as x approached infinity or negative infinity.  In general, if the $y$-values of a function approach a specific number as $x$ approaches infinity or negative infinity, we call that a horizontal asympote of the function.

\section{Horizontal Asymptotes}

\begin{definition}
  \mbox{}
  \begin{enumerate}
  \item The line $y=c$ is called a \emph{horizontal asymptote} of the graph of a function $y=f(x)$ if as $x \to -\infty$ or $x \to +\infty$, we have $f(x) \to c$.
  \end{enumerate}
\end{definition}

This means that as $x$ gets bigger and bigger (in the positive or negative direction), the $y$-values of the function get close to a particular value $y=c$.   Anther way to understand this that as $x$ gets bigger and bigger (in the positive or negative direction), the graph of the function approaches the horizontal line $y=c$.  

We know from looking at $f(x)=\frac{1}{x}$ that some rational functions have horizontal asymptotes.  The question we would like to investigate is which rational functions have horizontal asymptotes, and if a rational function does have a horizontal asymptote, how do we determine it's value?  Let's look at a few particular examples.

\begin{example} 

Consider the rational function $f(x) = \frac{x-1}{x-2}$. To intuitively understand $x\to +\infty$ intuitively, let's plug some big values for $x$:
    \begin{align*}
      f(100) &= \frac{99}{98} \approx 1.010... \\ f(1000) &= \frac{999}{998} \approx 1.001... \\ f(10000) &= \frac{9999}{9998} \approx 1.000...
    \end{align*}
It seems like $f(x) \to 1$ as $x \to +\infty$.  That would mean that the line $y=1$ is a horizontal asymptote for $f(x)$. 

The same strategy shows that $f(x) \to 1$ when $x \to -\infty$ as well, so that $y=1$ is the only horizontal asymptote for $f(x)$. %As for vertical asymptotes, we see that the only $x$-value for which $f(x)$ is undefined is $x=2$. So that's where we'll look, by choosing values for $x$ very close to $2$, but not equal to $2$. For example, we have that
%    \begin{align*}
%      f(2.1) &= \frac{1.1}{0.1} = 11 \\ f(2.01) &= \frac{1.01}{0.01} = 101 \\ f(2.001) &= \frac{1.001}{0.001} = 1001 
%    \end{align*}
%    indicates that $f(x) \to +\infty$ as $x \to 2^+$. Similarly, you can check that $f(x) \to -\infty$ as $x \to 2^-$, so the line $x=2$ is a vertical asymptote for $f(x)$ (in fact, the only one). 

Here's a graph of the function $f(x) = \frac{x-1}{x-2}$.  Notice how the function gets close to the line $y=1$ as it goes off the page on the left and right sides.

\begin{image}
\begin{tikzpicture}
    \begin{axis}
      \addplot[samples=200,domain=2.01:7]{(x-1)/(x-2)};
      \addplot[samples=200,domain=-7:1.99]{(x-1)/(x-2)};
 %     \addplot[dashed, color=penColor] coordinates {(2,-7) (2,7)};
      \addplot[dashed, color=penColor, domain=-7:7]{1};
    \end{axis}
\end{tikzpicture}
\end{image}

\end{example}

\begin{example}
Consider now the rational function $g(x) = \frac{x-1}{x^2-3x+2}$. Let's do as above, and start looking for horizontal asymptotes.
  \begin{align*}
   g(100) &\approx 0.010... \\
   g(1000) &\approx 0.001... \\
    g(10000) &\approx 0.000...
  \end{align*}
  This suggests that $g(x) \to 0$ as $x \to +\infty$. Similarly, you can convince yourself that $g(x) \to 0$ as $x \to -\infty$, so that $y=0$ is the only horizontal asymptote.  Here's the graph of $g(x) = \frac{x-1}{x^2-3x+2}$.  Notice how the function gets close to the line $y=0$ as it goes off the page on the left and right sides.

      \begin{image}
        \begin{tikzpicture}
          \begin{axis}
            \addplot[samples=200,domain=-7:0.99]{1/(x-2)};
            \addplot[samples=200,domain=1.01:1.99]{1/(x-2)};
            \addplot[samples=200,domain=2.01:7]{1/(x-2)};
            \draw[fill=white](axis cs:1,-1)circle(1mm);
       %     \addplot[dashed, thick, color=penColor] coordinates {(2,-7) (2,7)};
            \addplot[thick, dashed, color=penColor, domain=-7:7]{0};
          \end{axis}
        \end{tikzpicture}
      \end{image}

You may wonder why there is a hole in this graph!  We will explore that more in the next section.

\end{example}



%\begin{exploration}
%  A mathematical model for the population $P$, in thousands, of a certain species of bacteria, $t$ days after it is introduced to an environment, is given by $P(t) = \frac{200}{(7-t)^2}$, $0 \leq t < 7$.
%  \begin{enumerate}
%  \item Find and interprete $P(0)$.
%  \item When will the population reach $200,000$?
%  \item Determine the behavior of $P$ as $t \to 7^-$. Interpret this result graphically and within the context of the problem.
%  \end{enumerate}
%\end{exploration}

Now, you must be asking yourself if every time we want to test for vertical or horizontal asymptotes, we need to keep plugging values and guessing. Fortunately, the answer is ``no''.   With a little bit of algebra, it becomes much more apparent what the end behavior of a rational function will be.  Formal justificatives require Calculus --- we'll be content with getting intuition for now.

We will need the following theorem.

\begin{theorem}
Functions of the form $$f(x)=x^{-n}=\frac{1}{x^n}$$ where $n$ is a counting number, have horizontal asymptotes of $y=0$.  In particular, as $x \rightarrow \infty, y \rightarrow 0$ and   as $x \rightarrow -\infty, y \rightarrow 0$.
\end{theorem}

Hopefully this theorem seems intuitive.  If $f(x)=\frac{1}{x}$ goes to 0 as $x$ gets big, then raising $x$ to a higher power should just make the function get close to zero more quickly.  If we take this theorem as a given, we can use it and some algebra to rewrite other rational functions so that we can find their horizontal asymptotes.

\begin{example}  Find the horizontal asympotes of $f(x) = \frac{x-1}{x-2}$.

\begin{explanation}
Notice that this is the same function we looked at in an earlier example.  From plugging in points, we think this function should have a horizontal asymptote of $y=1$, meaning the $y$-values of this function get close to $1$ as $x$ gets big (either big and positive or big and negative).  Now, lets show this with some algebra.  The trick is going to be to \textbf{divide the top and bottom of the rational function by the highest power of $x$ in the denominator}.
\begin{align*}
f(x) =& \frac{x-1}{x-2}\\
=& \frac{\frac{1}{x}\left(x-1\right)}{\frac{1}{x}\left(x-2\right)} \\
=& \frac{\frac{1}{x} \cdot x-\frac{1}{x}}{\frac{1}{x}\cdot x-\frac{1}{x} \cdot 2} \text{ \, \,\,    distributing} \\
=& \frac{1-\frac{1}{x}}{1-\frac{2}{x}} \\
\end{align*}

Now, notice that we have $\frac{1}{x}$ and $\frac{2}{x}$ in our new way of writing this rational function, and we know that both of these functions approach $0$ as $x$ goes to infinity and negative infinity.  This allows us to say  that,
\[
\text{as } x \rightarrow \infty, \frac{1-\frac{1}{x}}{1-\frac{2}{x}} \rightarrow \frac{1-0}{1-0}=\frac{1}{1}=1
\]
and
\[
\text{as } x \rightarrow -\infty, \frac{1-\frac{1}{x}}{1-\frac{2}{x}} \rightarrow \frac{1-0}{1-0}=\frac{1}{1}=1
\]

This tells us that $y=1$ is a horizontal asymptote of $f$.
\end{explanation}
\end{example}

The technical reason why this works require calculus, but for now it is enough to know that if you can rewrite a rational function so it contains terms of $\frac{1}{x^n}$, then as $x$ goes to infinity or negative infinity, those terms will become $0$.  Let's try another example.

\begin{example}  Find the horizontal asympotes of $f(x) = \frac{5x^3-3x-1}{2x^3-2x^2+8}$.

\begin{explanation}
Recall that we want to \textbf{divide the top and bottom of the rational function by the highest power of $x$ in the denominator}.  For this function, we want to divide by $x^3$.
\begin{align*}
f(x) =& \frac{5x^3-3x-1}{2x^3-2x^2+8}\\
=& \frac{\frac{1}{x^3}\left(5x^3-3x-1\right)}{\frac{1}{x^3}\left(2x^3-2x^2+8\right)} \\
=& \frac{\frac{1}{x^3} \cdot 5x^3-\frac{1}{x^3}\cdot 3x - \frac{1}{x^3}}{\frac{1}{x^3}\cdot 2x^3-\frac{1}{x^3} \cdot 2x^2 + \frac{1}{x^3} \cdot 8} \text{ \, \,\,    distributing} \\
=& \frac{5-\frac{3}{x^2}-\frac{1}{x^3}}{2-\frac{2}{x}+\frac{8}{x^3}} \\
\end{align*}

Now, notice that we have $\frac{3}{x^2}$, $\frac{1}{x^3}$, $\frac{2}{x}$ and $\frac{8}{x^3}$ in our new way of writing this rational function which all approach $0$ as $x$ goes to infinity and negative infinity.  This allows us to say  that,
\[
\text{as } x \rightarrow \infty, \frac{5-\frac{3}{x^2}-\frac{1}{x^3}}{2-\frac{2}{x}+\frac{8}{x^3}} \rightarrow \frac{5-0-0}{2-0+0}=\frac{5}{2}
\]
and
\[
\text{as } x \rightarrow -\infty, \frac{5-\frac{3}{x^2}-\frac{1}{x^3}}{2-\frac{2}{x}+\frac{8}{x^3}} \rightarrow \frac{5-0-0}{2-0+0}=\frac{5}{2}
\]

This tells us that $y=\frac{5}{2}$ is a horizontal asymptote of $f$.
\end{explanation}
\end{example}

You may have noticed a pattern.  In both examples above, we just ended up with the coefficients of the leading terms from the top and bottom of the rational function.  This pattern holds whenever the top and bottom fractions have the same degree.  The theorem below gives the pattern for horizontal functions in general.

\begin{theorem}
  {\bf Locating Horizontal Asymptotes:} Assume that $f(x) = p(x)/q(x)$ is a rational function, and that the leading coefficients of $p(x)$ and $q(x)$ are $a$ and $b$, respectively.
  \begin{itemize}
  \item If the degree of $p(x)$ is the same as the degree of $q(x)$, then $y=a/b$ is the unique horizontal asymptote of the graph of $y=f(x)$.
  \item If the degree of $p(x)$ is less than the degree of $q(x)$, then $y=0$ is the unique horizontal asymptote of the graph of $y=f(x)$.
  \item If the degree of $p(x)$ is greater than the degree of $q(x)$, then the graph of $y=f(x)$ has no horizontal asymptotes.
  \end{itemize}
\end{theorem}

The above theorem essentially says that one can detect horizontal asymptotes by looking at degrees and leading coefficients. Only the leading terms of $p(x)$ and $q(x)$ matter, and it makes no difference whether one considers $x\to +\infty$ or $x\to -\infty$. For example, how would you apply this to study horizontal asymptotes for the following function? $$ f(x) =  \frac{3x^6-5x^4+3x^3-3x^2 + 10x + 1}{5x^6 + 10000x^5 - 5x+2}$$

\begin{explanation}
  The degree of the numerator $p(x) = 3x^6-5x^4+3x^3-3x^2 + 10x + 1$ and the degree of the denominator $q(x) = 5x^6 + 10000x^5 - 5x+2$ are equal, namely, to $6$. Thus, since the leading coefficient of $p(x)$ is $3$ and the leading coefficient of $q(x)$ is $5$, we conclude that the line $y=3/5$ is the only horizontal asymptote for this rational function.
\end{explanation}


\section{Slant asymptotes}

From the theorem above, we can see that in the third case when the degree of the numerator is bigger than the degree of the denominator, there are no horizontal asymptotes..  It would be natural to ask whether we can still say anything about the rational function in that case.  It turns out that we can, and the key is to use \textbf{long division of polynomials}.  Whenever the degree of the numerator is bigger that the degree of the denominator, we can use long division to rewrite the function as a polynomials plus a rational function where the degree of the numerator is \textit{less} than the degree of the denominator. 

\begin{example}
For example, lets consider the rational function $$h(x) = \frac{x^3+3x^2+x+1}{x^2+1}.$$  In the section on polynomial long division, we determined that that $f(x) = x^3+3x^2+x+1$ divided by $g(x) = x^2+1$ would give us 
$$  \frac{x^3+3x^2+x+1}{x^2+1} = x+3 - \frac{2}{x^2+1}.  $$
This means we have rewritten $h(x) = \frac{x^3+3x^2+x+1}{x^2+1}$ as the sum of the polynomial $x+3$ and the rational function $\frac{2}{x^2+1}$.  By the previous therem, we can say that
\[
\text{as } x \rightarrow \pm \infty, \frac{2}{x^2+1} \rightarrow 0
\]
This implies that when $x$ is really big, $h(x) = \frac{x^3+3x^2+x+1}{x^2+1}$ behaves very similarly to $x+3$ (because the rest of the function essentially becomes very tiny).   Recall that when a graph approaches a line, we call that line an asymptote.  We call the line $y=x+3$ a \textbf{slant asymptote} of the function $h$ because, unlike a vertical or horizontal line, the line $y=x+3$ is slanted.
\end{example}

\begin{definition}
  The line $y=mx+b$, where $m\neq 0$, is called a \emph{slant asymptote} of the graph of a function $y=f(x)$ if as $x \to -\infty$ or as $x\to +\infty$, we have $f(x) \to mx+b$.
\end{definition}

Note that saying that $y=mx+b$ is a slant asymptote for the graph of $y=f(x)$ is the same thing as saying that $y=0$ is a horizontal asymptote for the graph of the difference function $y=f(x)-(mx+b)$.




\begin{example} Do the following rational functions have slant asymptotes? If so, what are their line equations?
  \begin{enumerate}
  \item $f(x) = \frac{x^2-4x+2}{1-x}$. \\[1em]
    \begin{explanation}
      When trying to find slant asymptotes, long division is the way to go. Performing it, we see that $$    \frac{x^2-4x+2}{1-x} = -x+3 - \frac{1}{1-x}.  $$Since $1/(1-x) \to 0$ as $x \to \infty$ or $x \to -\infty$ and $y=-x+3$ describes a line, we conclude that $y=-x+3$ is a slant asymptote for the graph of $y=f(x)$.
  \begin{image}
 \begin{tikzpicture}
     \begin{axis}
       \addplot[samples=200,domain=1.01:7]{(x^2-4*x+2)/(1-x)};
       \addplot[samples=200,domain=-7:0.99]{(x^2-4*x+2)/(1-x)};
       \addplot[dashed,thick,domain=-7:7, color=penColor]{-x+3};
     \end{axis}
 \end{tikzpicture}
 \end{image}
    \end{explanation}
    
  \item $f(x) = \frac{x^2-4}{x-2}$. \\[1em]
    \begin{explanation}
      We may just simplify it to $f(x) = x+2$, valid for all $x \neq 2$. We may regard this as a long division for which the remainder is zero, as in $$  f(x) = x+2+\frac{0}{x-2},  $$and since $0/(x-2) \to 0$ as $x\to +\infty$ or $x\to -\infty$, it follows that $y=x+2$ is a slant asymptote for the graph of $y=f(x)$, even though the graph is just said line with a hole!
      \begin{image}
        \begin{tikzpicture}
          \begin{axis}
            \addplot[domain=-7:7]{x+2};
           \draw[fill=white](axis cs:2,4)circle(1mm);
          \end{axis}
        \end{tikzpicture}
      \end{image}      
    \end{explanation}
  \item $f(x)=\frac{x^3-4x^2+5x-1}{x-1}$. \\[1em]
    \begin{explanation}
      Performing the long division, as before, we see that $$   f(x) = \frac{x^3-4x^2+5x-1}{x-1} = x^2-3x+2 + \frac{1}{x-1}.  $$As expected, $1/(x-1) \to 0$ when $x \to +\infty$ or $x\to -\infty$, but $y=x^2-3x+2$ is not a line equation. Hence there are no slant asymptotes for the graph of $f(x)$.
      \begin{image}
        \begin{tikzpicture}
          \begin{axis}
            \addplot[samples=200,domain=1.01:7]{(x^3-4*x^2+5*x-1)/(x-1)};
            \addplot[samples=200,domain=-7:0.99]{(x^3-4*x^2+5*x-1)/(x-1)};
            \addplot[dashed,thick,smooth,domain=-7:7,color=penColor]{x^2-3*x+2};
          \end{axis}
        \end{tikzpicture}
      \end{image}
    \end{explanation}
Notice that even though $x^2-3x+2$ is not a slant asymptote (because it is not a straight line), we can see from the graph that it is asymptotic to a parabola.  This means that the parabola still tells us about the end behavior of $f$.  Since $x^2-3x+2$ is an even degree polynomial with a positive leading coefficient, we know that 
\[
\text{as } x \rightarrow \pm \infty, x^2-3x+2 \rightarrow \infty.
\]
 Because of this, we can also say that 
\[
\text{as } x \rightarrow \pm \infty, \frac{x^3-4x^2+5x-1}{x-1} \rightarrow \infty.
\]

  \item $f(x) = \frac{x^3+2x^2+x+1}{x^3-2x^2-x+1}$. \\[1em]
   \begin{explanation}
      Long division shows that: $$   f(x) = \frac{x^3+2x^2+x+1}{x^3-2x^2-x+1} = 1 + \underbrace{\frac{4x^2+2x}{x^3-2x^2-x+1}}_{\to 0}.  $$The indicated remainder goes to zero when $x \to +\infty$ or $x\to -\infty$ simply because the degree of the numerator is lower than the degree of the numerator. The remaining quotient does give us the asymptote $y=1$. But this is not a slant asymptote, it is a horizontal asymptote (as you might have expected).
%            \begin{image}
%        \begin{tikzpicture}
%          \begin{axis}
%            \addplot[samples=300,domain=-7:-0.83]{(x^3+2*x^2+x+1)/(x^3-2*x^2-x+1)};
%            \addplot[samples=300,domain=-0.8:0.554]{(x^3+2*x^2+x+1)/(x^3-2*x^2-x+1)};
%             \addplot[samples=300,domain=0.556:2.24]{(x^3+2*x^2+x+1)/(x^3-2*x^2-x+1)};
%             \addplot[samples=300,domain=2.26:7]{(x^3+2*x^2+x+1)/(x^3-2*x^2-x+1)};
%             \addplot[dashed,thick,domain=-7:7, color=penColor]{1};
%          \end{axis}
%        \end{tikzpicture}
%      \end{image}
      Note that asymptotes are really concerned about the end behavior of the function. In the above example, the line $y=1$ does intersect the graph of $y=f(x)$, but this is fine --- the graph still only approaches said line as $x \to +\infty$ and $x \to -\infty$.
   \end{explanation}
  \item $f(x) = \frac{x^2+1}{x^5-4}$. \\[1em]
    \begin{explanation}
      Since the degree of the numerator is smaller than the degree of the denominator, you can think of the long division as already having been performed, as in $$ f(x) = \frac{x^2+1}{x^5-4} = 0 + \frac{x^2+1}{x^5-4}.  $$Again, the remainder goes to zero when $x \to +\infty$ or $x\to -\infty$. So, the line $y=0$ would be an asymptote, but it is horizontal, not slant, as in the previous item.

  \begin{image}
 \begin{tikzpicture}
     \begin{axis}
       \addplot[samples=200,domain=-7:1.318]{(x^2+1)/(x^5-4)};
       \addplot[samples=200,domain=1.32:7]{(x^2+1)/(x^5-4)};
       \addplot[dashed,thick,domain=-7:7, color=penColor]{0};
     \end{axis}
 \end{tikzpicture}
 \end{image}

      
    \end{explanation}
  \end{enumerate}

\end{example}

The above examples suggest that if the degree of the numerator is at least two higher than the degree of the numerator, what survives outside the remainder has degree higher than one, and thus does not describe a line equation --- meaning no slant asymptotes. Similarly, if the degree of the numerator is equal or lower to the degree of the denominator, there's ``not enough quotient left" to describe a line equation. This is not a coincidence, but a general fact.

\begin{callout}
  {\bf Theorem (on slant asymptotes):} Let $f(x) = p(x)/q(x)$ be a rational function for which the degree of $p(x)$ is exactly one higher than the degree of $q(x)$. Then the graph of $y=f(x)$ has the slant asymptote $y=L(x)$, where $L(x)$ is the quotient obtained by dividing $p(x)$ by $q(x)$. If the degree of $p(x)$ is not exactly one higher than the degree of $q(x)$, there is no slant asymptote whatsoever.
\end{callout}

  Unlike what happened for horizontal and vertical asymptotes, the above theorem does not \emph{immediately} tell you what is the line equation describing the slant asymptote. We must resort to long division.

%\typeout{************************************************}
%\typeout{Summary}
%\typeout{************************************************}

\begin{summary}\begin{itemize}
%\item A \emph{rational} function is, as the name suggests, a function defined as a \emph{ratio} $f(x) = p(x)/q(x)$ of two polynomials $p(x)$ and $q(x)$. It makes sense for all real values of $x$ \emph{except} for those such that $q(x) = 0$, as one cannot divide by zero.
\item There are three types of asymptotes for rational functions: vertical asymptotes, horizontal asymptotes, and slant asymptotes. The latter occurs when the degree of the numerator is exactly one higher than the degree of the denominator.
\end{itemize}\end{summary}



\end{document}
