\documentclass{ximera}

\input{../../preamble.tex}

\author{Elizabeth Campolongo}
\acknowledgement{https://www.stitz-zeager.com/szprecalculus07042013.pdf}

\begin{document}
\begin{exercise}
For the following word problems, determine what type of function is needed to solve them (e.g., linear, quadratic, cubic, logarithmic, exponential, etc.) and give the function. 
%
\begin{enumerate}
%
\item Mary has decided to take up alpaca farming to make home-spun yarn, thus feeding her other hobby of sustainable knitting. She just inherited a large parcel of land in Ashtabula, which is perfect for her to make this dream come true! Her funds are still limited, so she can afford only enough fencing material to build a rectangular pasture with 200 linear feet of fence. She will construct it adjacent to a conveniently placed stream (removing the requirement for fencing on this side). 
%ex2.3.4 in stitz-zeager

She wants to maximize the area of the pasture. What type of function will she need to do this calculation?\\
$\wordChoice{\choice{linear}\choice[correct]{quadratic}\choice{cubic}\choice{logarithmic}\choice{exponential}}$
\begin{exercise}
She will need a quadratic equation to maximize the Area. If we let $\ell$ denote the length adjacent to the river and $w$ denote the width of the pasture, the formula for the area in terms of width is
$$A(w) = \answer{-2}w^2 + \answer{200w}.$$

\begin{exercise} %ex20 of same section
Mary starts to worry that leaving the alpaca along the stream could be dangerous with the storms that have been rolling in recently, and she decides to construct a freestanding pasture. She still only has 200 feet worth of fencing; what is the new function to maximize the area in terms of width?\\
$A(w) = \answer{-}w^2 + \answer{100w}$.
\end{exercise}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%
\item  Iodine-131 is a commonly used radioactive isotope used to help detect how well
the thyroid is functioning. Suppose the decay of Iodine-131 follows the model $A(t) = A_0e^{kt}$,
and that the half-life10 of Iodine-131 is approximately 8 days. 
If 5 grams of Iodine-131 is present
initially, find a function which gives the amount of Iodine-131, $A$, in grams, $t$ days later.

What type of equation would be needed to solve this?\\
$\wordChoice{\choice{linear}\choice{quadratic}\choice{cubic}\choice{logarithmic}\choice[correct]{exponential}}$.
\begin{exercise}
The model is $A(t) = A_0e^{kt}$, where $A(0)=A_0$ the initial amount of the element, $k<0$. is the constant of proportionality which satisfies the equation. \\
Now, find the function to represent the amount of Iodine-131, in grams, after $t$ days. \\
$A(t) = \answer{5}e^{-\answer{t} \frac{\answer{\ln(2)}}{\answer{8}}}$
\end{exercise}


%%%%%%%%%%%%%%%%%%%%%%
\item The half-life of the radioactive isotope Carbon-14 is about 5730 years.

Express the amount of Carbon-14 left from an initial $N$ milligrams
as a function of time $t$ in years.

What type of equation would be needed to solve this?\\
$\wordChoice{\choice{linear}\choice{quadratic}\choice{cubic}\choice{logarithmic}\choice[correct]{exponential}}$.
\begin{exercise}
The model is $A(t) = A_0e^{kt}$, where $A(0)=A_0$ the initial amount of the element, $k<0$. is the constant of proportionality which satisfies the equation. \\
Now, find the function to represent the amount of Carbon-14, in grams, after $t$ years. \\
$A(t) = \answer{N}\answer{e}^{-\answer{t} \frac{\answer{\ln(2)}}{\answer{5730}}}$
\end{exercise}



%%%%%%%%%%%%%%%%%%%%%%
\item To mail a rectangular box with USPS Parcel Post, the box must satisfy the inequality ``Length $+$ Girth $\leq 130$ inches." Assuming we want to ship a rectangular box with a square face of side-length $x$. The length is the longest side of the box (not the square part). The girth is the sum of the four sides of the square ($4x$).

We are looking to maximize the volume of such a box. What type of equation would be needed to solve this?\\
$\wordChoice{\choice{linear}\choice{quadratic}\choice[correct]{cubic}\choice{logarithmic}\choice{exponential}}$.
%
\begin{exercise}
We maximize the volume with a cubic equation given by $V(x) = \answer{-4}x^3+\answer{130x^2}$.
\begin{exercise}
If we instead wanted to ship via another method where the ``Length $+$ Girth $\leq 130$ inches," what would be our new volume function for the same square-faced rectangular box?\\
We maximize the volume with a cubic equation given by $V(x) = \answer{-4}x^3+\answer{108x^2}$.
\end{exercise}
\end{exercise}

%%%%%%%%%%%%%%%%%%%%%%
\item The information entropy $H$, in bits, of a randomly generated password consisting
of $n$ characters is given by $H = n\log_2(N)$, where $N$ is the number of possible symbols for each
character in the password. In general, the higher the entropy, the stronger the password.

If a 7 character case-sensitive password is comprised of letters and numbers only, find the
associated information entropy.
What type of equation would be needed to solve this?\\
$\wordChoice{\choice{linear}\choice{quadratic}\choice{cubic}\choice[correct]{logarithmic}\choice{exponential}}$.
\begin{exercise}
Write out the equation to solve this question.\\
$H = \answer{7}\log_{\answer{2}}(\answer{62}) = \answer{7}\frac{\answer{\ln(62)}}{\ln(\answer{2})}$.

\begin{exercise}
How many possible symbol options per character is required to produce a 7 character password
with an information entropy of 50 bits?\\
Again, write out the equation to solve this question.\\
$\answer{50} = \answer{7}\log_{\answer{2}}(\answer{N})$.
\end{exercise}
\end{exercise}


%%%%%%%%%%%%%%%%%%%%%%%
\item Suppose you invest \$2000 in an account which offers 7.125\% compounded monthly. Without adding to your initial investment, you want to know how much will be in the account after $t$ years. 

What type of equation would be needed to solve this?\\
$\wordChoice{\choice{linear}\choice{quadratic}\choice{cubic}\choice{logarithmic}\choice[correct]{exponential}}$.
\begin{exercise}
Compound interest can be represented by the function
$$A(t) = P\left(1+\frac{r}{n}\right)^{nt},$$
where $P$ is the principal invested at an annual rate $r$, and the interest is compounded $n$ times per year. Then $A(t)$ represents the amount in the account after
%
$\wordChoice{\choice{n \ months}\choice{t \ months}\choice{n \ years}\choice[correct]{t \ years}}$.

\begin{exercise}
Express the amount in the account $A$ as a function of the term of the investment in $t$ years.\\
$A(t) = \answer{2000}(\answer{1}+ \answer{\frac{0.07125}{12}})^{\answer{12t}}$
\end{exercise}
\end{exercise}


\end{enumerate}

\end{exercise}
\end{document}